If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-13 GroupPolarizationStatmod/paper/intro-latest.tex
revised, expanded, focused intro-latest

Replication is not enough:
strong, mechanistic models must be developed to be able to predict Type I and II
error rates *ahead of time*, as we demonstrate here. We only demonstrate how to
calculate expected Type I error rates for a measurement design, i.e., the rate of false 
positives. Low statistical power exacerbates false detection rates. So, to be confident in replication,
a pre-registered measurement design should use this style of simulation to 
estimate the statistical power. Tangled, weak, informal
verbal models of complex psychological and social phenomena like those that motivated
empirical studies of group polarization tend to be vulnerable
to stastical weaknesses. It is difficult to incorporate prior knowledge such
as experiment structure without a sufficiently rigorous and appropriate 
formal system \cite{Smaldino2017,Smaldino2020,Flake2020,Deffner2024}.

% This work is motivated by a core conviction that better thinking leads
% to better science. Theoretical clarity, careful measurement, and
% principled statistical modeling are not academic luxuries, but essential
% tools for empirical science. Theoretical and statistical confusions in
% the group polarization literature have lain dormant, undermining
% measurements of opinion change in group polarization unbeknownst to
% generations of researchers (Brown 1986).

\subsection{Identifiability and the Illusion of
Change}\label{identifiability-and-the-illusion-of-change}

\begin{itemize}
\tightlist
\item
  Many group polarization studies report opinion change based on
  pre/post differences, but assume a transparent mapping between
  observed scores and latent beliefs.
\item
  Yet ordinal opinion data can mask deeper structural ambiguities that
  invalidate these assumptions.
\item
  In what follows, we first describe the basic structure of group
  polarization experiments, then explain how ordinal measurement can
  lead to illusory shifts.
\item
  Finally, we introduce a null-consistency modeling procedure that
  operationalizes this insight, and show how it informs the credibility
  of reported effects.
\end{itemize}

\subsubsection{Experimental structure of group polarization
studies}\label{experimental-structure-of-group-polarization-studies}

\begin{itemize}
\tightlist
\item
  Most group polarization experiments follow a familiar structure:
  measure individual opinions, allow group discussion, then re-measure.
\item
  The central claim is that like-minded individuals become more extreme
  after deliberation.
\item
  Crucially, opinions are typically measured with Likert-style ordinal
  scales (e.g., from ``Strongly disagree'' to ``Strongly agree'').
\item
  Analyses often treat these as interval-scaled, calculating mean
  differences and testing for significance.
\end{itemize}

\subsubsection{The ordinal-continuous mismatch
problem}\label{the-ordinal-continuous-mismatch-problem}

\begin{itemize}
\tightlist
\item
  Ordinal scores do not have a fixed interval structure --- they
  represent ranked categories, not true distances.
\item
  A given change in score may reflect a shift in opinion, increased
  certainty, or both.
\item
  For example, if opinions become more tightly clustered (lower
  variance) post-deliberation, the group mean can appear to shift even
  when the latent mean remains unchanged.
\item
  This opens the door to \emph{null-consistent detections} --- apparent
  effects that emerge from measurement artifacts rather than real
  opinion movement.
\end{itemize}

\subsubsection{Description of the null-consistency modeling
procedure}\label{description-of-the-null-consistency-modeling-procedure}

\begin{itemize}
\tightlist
\item
  We formalize this concern using a generative model that holds the
  latent opinion mean constant while allowing variance to shrink.
\item
  The goal is to test whether an observed group polarization result
  could arise from such a stable-mean, shifting-variance process.
\item
  If this model can reproduce the pre/post measurement pattern, the
  observed result is non-diagnostic --- it is consistent with a null
  hypothesis of no true change.
\item
  This is a form of constructive falsification: it doesn't prove the
  effect is false, but shows it is not uniquely explained by change.
\end{itemize}

\subsubsection{Implications for statistical power and detection
credibility}\label{implications-for-statistical-power-and-detection-credibility}

\begin{itemize}
\tightlist
\item
  Even if an effect is real, detecting it reliably requires statistical
  power --- the ability to distinguish true shifts from noise or
  artifacts.
\item
  Studies using ordinal measures and small samples may be especially
  vulnerable to false positives under the null.
\item
  To assess this, we fit Bayesian ordered probit models to simulated
  data and estimate the power to detect effects of various sizes.
\end{itemize}

\subsubsection{Transition to empirical review and FDR
estimation}\label{transition-to-empirical-review-and-fdr-estimation}

\begin{itemize}
\tightlist
\item
  Having established a model-based definition of null-consistent
  detection, we now apply it to published group polarization
  experiments.
\item
  By combining this logic with simulations seeded from published data,
  we estimate false discovery rates across conditions.
\item
  This isn't a judgment about whether past detections were true. It's a
  demonstration of how to estimate false detection rates
  properly---using models that actually fit the structure of the data.
  We calculate FWER and FDR for the original study designs at three
  effect size levels---small, medium, and large---based on Cohen's
  guidelines (Cohen 1988, Ch. 2.2.3) and Bayesian methods from Liddell
  and Kruschke (Liddell and Kruschke 2018). The verdict is clear: even
  if these studies had used better models, most still wouldn't have had
  the power to rule out false positives. That makes replication a weak
  signal of reliability. If future group polarization studies want to
  produce meaningful result need stronger designs---and that starts with
  a clean break from outdated statistical habits.
\end{itemize}

The null-consistency procedure functions as a test of model
indistinguishability: it demonstrates that a reported group polarization
effect could plausibly arise from a stable-latent process combined with
changing opinion precision. This logic echoes foundational concerns in
causal inference, where two competing models---one causal, one not---can
imply the same observed data distribution and are therefore empirically
indistinguishable without further assumptions (Pearl 2000; Spirtes,
Glymour, and Scheines 2000). In such cases, inference from observation
to mechanism is invalid. Similarly, in psychometrics, it is well known
that different configurations of latent traits and response thresholds
can yield identical observed ordinal patterns, making the true structure
unidentifiable without additional constraints (Borsboom 2005). The
present approach does not attempt to estimate the probability of false
detection, but rather shows that under ordinal measurement, a null model
can reproduce the observed pattern---rendering the result non-diagnostic
of a true latent shift. This establishes a stringent standard: an effect
must not only be detected, but shown to be incompatible with plausible
null-generating mechanisms.

>>>>

Replication is not enough:
strong, mechanistic models must be developed to be able to predict Type I and II
error rates \emph{ahead of time}, as we demonstrate here. We only demonstrate how to
calculate expected Type I error rates for a measurement design, i.e., the rate of false 
positives. Low statistical power exacerbates false detection rates. So, to be confident in replication,
a pre-registered measurement design should use this style of simulation to 
estimate the statistical power. Tangled, weak, informal
verbal models of complex psychological and social phenomena like those that motivated
empirical studies of group polarization tend to be vulnerable
to stastical weaknesses. It is difficult to incorporate prior knowledge such
as experiment structure without a sufficiently rigorous and appropriate 
formal system \cite{Smaldino2017,Smaldino2020,Flake2020,Deffner2024}.

\subsection{Group polarization research}

(DO THE MINIMUM AFTER EVERYTHING ELSE)

To situate this current paper it will help to introduce some relevant research
traditions and common experimental designs in group polarization, using the ten
studies included in our corpus as touchstones. To begin, *group polarization*
is defined as...(continue basic definition with some examples from our corpus).

Paragraph dos: 
Common experimental, measurement, and statistical procedures to induce and
detect group polarization

Paragraph tres: How this procedure has been used in the great pointless theoretical milieu
that raged for decades with apparently no strong determination on the outcome
or even what the fight was all about.

Paragraph quatro e final: our corpus.

\subsection{Overview}

In the next section, we introduce the various formal and computational Methods 
used to obtain our findings outlined above. We then explain our findings in detail in the
Analysis section. We close with a Discussion of how this is not a defeat, but
a new beginning for group polariation, given that there are good reasons to expect
group polarization to occur. We also explain how to expanding this analysis to
examine other published studies on group polarization and across social and behavioral
science disciplines.
